\documentclass[11pt,a4paper]{article}

%% ── Packages ─────────────────────────────────────────────────────────
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{tikz}
\usepackage{enumitem}
\usepackage[margin=2.5cm]{geometry}
\usepackage{xcolor}
\usepackage{algorithm}
\usepackage{algpseudocode}

\usetikzlibrary{matrix,positioning,arrows.meta}

%% ── Theorem-like environments ────────────────────────────────────────
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

%% ── Convenience macros ──────────────────────────────────────────────
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\N}{\mathbb{N}}
\DeclareMathOperator{\modd}{mod}

%% ── Title ───────────────────────────────────────────────────────────
\title{\textbf{The Alien Tiles Problem}\\[4pt]
       \large CSPLib Problem 027 --- Problem Description,\\
       ILP and SAT Models, and Datasets}
\author{}
\date{\today}

%% =====================================================================
\begin{document}
\maketitle

\begin{abstract}
The \emph{Alien Tiles} puzzle is a combinatorial puzzle played on a
square grid of coloured tiles.  A click on any tile rotates the colour
of that tile and every other tile sharing its row or column.  This
document gives a formal description of the problem (CSPLib
prob027~\cite{csplib027}), works through a concrete example, states the
key constraints, and formulates the problem as both an Integer Linear
Program (ILP) and a Boolean Satisfiability (SAT) instance.  Benchmark
datasets are described and cited.
\end{abstract}

\tableofcontents
\bigskip

%% =====================================================================
\section{Introduction}\label{sec:intro}

The Alien Tiles puzzle was originally available for play at
\url{http://www.alientiles.com}~\cite{gent2000symmetry}.  It was listed
as Problem~027 in \emph{CSPLib}, a library of test problems for
constraint solvers~\cite{csplib027}.  The puzzle belongs to the broader
family of ``Lights Out'' games and is closely related to problems in
coding theory over finite
fields~\cite{chlond2006alien,gent2000symmetry}.

The central combinatorial question is: \emph{given an $N\times N$ grid
with $c$ colours and a goal state, what is the minimum number of clicks
required to reach that state from the all-zero start state?}  A harder
meta-question asks for the \emph{hardest} solvable instance, i.e., the
goal state whose minimum-click solution is largest.

%% =====================================================================
\section{Problem Description}\label{sec:description}

\begin{definition}[Alien Tiles Puzzle]\label{def:puzzle}
An instance of the Alien Tiles puzzle is specified by a triple
$(N, c, T)$ where
\begin{itemize}[nosep]
  \item $N \in \N$ is the grid size (the grid has $N\times N$ cells),
  \item $c \in \N,\; c \ge 2$ is the number of colours (equivalently,
        the modulus), and
  \item $T \in \Z_c^{N\times N}$ is the \emph{target} (goal) state,
        where $\Z_c = \{0,1,\dots,c-1\}$.
\end{itemize}
The \emph{start state} is the all-zero matrix $S = \mathbf{0}$.
\end{definition}

\paragraph{Click mechanics.}
A \emph{click} on cell $(i,j)$ adds~$1$ (modulo~$c$) to the colour of
every cell that shares row~$i$ or column~$j$ with $(i,j)$.  Formally,
clicking $(i,j)$ modifies the state matrix $G$ as follows:
\[
  G'_{r,k} =
  \begin{cases}
    (G_{r,k} + 1) \bmod c & \text{if } r = i \text{ or } k = j,\\[2pt]
    G_{r,k} & \text{otherwise}.
  \end{cases}
\]
Because clicks are \emph{commutative} (the order does not matter), a
solution is fully characterised by a \emph{click matrix}
$X = (x_{i,j}) \in \Z_c^{N\times N}$, where $x_{i,j}$ is the number
of times cell $(i,j)$ is clicked.  Any $c$ clicks on the same cell
produce no net effect, so $0 \le x_{i,j} \le c-1$.

\paragraph{Effect equation.}
After applying all clicks, the resulting colour of cell $(r,k)$ is
\begin{equation}\label{eq:effect}
  G_{r,k} = \left(\sum_{j=1}^{N} x_{r,j}
             + \sum_{i=1}^{N} x_{i,k}
             - x_{r,k}\right) \bmod c.
\end{equation}
Note that $x_{r,k}$ is counted twice (once in row~$r$, once in
column~$k$), hence the subtraction of $x_{r,k}$.

\paragraph{Feasibility.}
A click matrix $X$ is a \emph{solution} if and only if
\begin{equation}\label{eq:feasibility}
  \forall\, r,k \in \{1,\dots,N\}:\quad
  \left(\sum_{j=1}^{N} x_{r,j} + \sum_{i=1}^{N} x_{i,k}
        - x_{r,k}\right) \bmod c = t_{r,k},
\end{equation}
where $t_{r,k}$ is the target colour of cell $(r,k)$.

%% =====================================================================
\section{Illustrative Example}\label{sec:example}

\begin{example}[$4\times 4$ grid, $c=3$ colours]\label{ex:4x4}
Consider $N=4$ and $c=3$ (colours 0, 1, 2 cycling as
$\text{Red}\to\text{Green}\to\text{Blue}\to\text{Red}$).

\medskip
\noindent\textbf{Step 1: A single click.}\;
Suppose we click cell $(1,1)$ once.  Every cell in row~1 and column~1
has its colour increased by~1 (mod~3).  The resulting state is:

\begin{center}
\begin{tikzpicture}[
    cell/.style={minimum size=8mm, draw, thick, font=\small},
    highlight/.style={cell, fill=green!30},
    normal/.style={cell, fill=gray!10},
  ]
  \matrix (m) [matrix of nodes, nodes in empty cells,
    row sep=-\pgflinewidth, column sep=-\pgflinewidth]{
    |[highlight]| 1 & |[highlight]| 1 & |[highlight]| 1 & |[highlight]| 1 \\
    |[highlight]| 1 & |[normal]|   0 & |[normal]|   0 & |[normal]|   0 \\
    |[highlight]| 1 & |[normal]|   0 & |[normal]|   0 & |[normal]|   0 \\
    |[highlight]| 1 & |[normal]|   0 & |[normal]|   0 & |[normal]|   0 \\
  };
  \node[above=2mm of m-1-1] {\footnotesize $j{=}1$};
  \node[above=2mm of m-1-2] {\footnotesize $j{=}2$};
  \node[above=2mm of m-1-3] {\footnotesize $j{=}3$};
  \node[above=2mm of m-1-4] {\footnotesize $j{=}4$};
  \node[left=2mm of m-1-1]  {\footnotesize $i{=}1$};
  \node[left=2mm of m-2-1]  {\footnotesize $i{=}2$};
  \node[left=2mm of m-3-1]  {\footnotesize $i{=}3$};
  \node[left=2mm of m-4-1]  {\footnotesize $i{=}4$};
\end{tikzpicture}
\end{center}

Cell $(1,1)$ has value~$1$ because it lies in both row~1 and column~1,
but the click only adds~1 once (not twice) to the clicked cell itself.

\medskip
\noindent\textbf{Step 2: Equivalence of solutions.}\;
Consider clicking once in \emph{every} cell of row~1 and twice in every
cell of row~2.  For cells in rows~3 and~4, each receives $4 \times 1 +
4 \times 2 = 12$ column-clicks from rows~1 and~2.  Since
$12 \bmod 3 = 0$, these cells are unchanged.  For cells in rows~1
and~2, the row-clicks further accumulate to multiples of~3.  The net
effect on the entire grid is zero---this is a \emph{null click pattern}.
Any solution $X$ can be replaced by $X' = X + \Delta$ (mod~$c$) where
$\Delta$ is such a null pattern, yielding an equivalent solution.

\medskip
\noindent\textbf{Step 3: Hardest instance.}\;
Among all solvable target states, we seek the one whose
\emph{minimum-click} representative (over all equivalent solutions) has
the \emph{largest} total click count.  For $N=4,\, c=3$, the answer is
$10$~clicks, and there are $19$ essentially different hardest instances
(up to symmetry)~\cite{gent2000symmetry}.
\end{example}

%% =====================================================================
\section{Key Constraints}\label{sec:constraints}

The structure of the Alien Tiles problem gives rise to several important
constraints and properties:

\begin{enumerate}[label=\textbf{C\arabic*.},leftmargin=*]
  \item \textbf{Commutativity.}\;
    The final state depends only on the \emph{multiset} of clicks, not
    their order.  Hence a solution is a matrix
    $X \in \Z_c^{N\times N}$, not a sequence.

  \item \textbf{Modular arithmetic.}\;
    All computations are performed in $\Z_c$.  In particular, $c$
    clicks on any single cell produce no net effect, so
    $x_{i,j} \in \{0, 1, \dots, c-1\}$.

  \item \textbf{Effect equation.}\;
    For each cell $(r,k)$, the constraint
    \[
      \Bigl(\sum_{j=1}^{N} x_{r,j}
            + \sum_{i=1}^{N} x_{i,k}
            - x_{r,k}\Bigr) \bmod c = t_{r,k}
    \]
    must hold.  This yields $N^2$ modular linear equations in $N^2$
    unknowns.

  \item \textbf{Equivalence classes.}\;
    Two click matrices $X$ and $X'$ are \emph{equivalent} if they
    produce the same goal state.  Null patterns (click matrices that
    produce the all-zero effect) form a subgroup of $\Z_c^{N\times N}$.

  \item \textbf{Symmetry.}\;
    The problem exhibits a symmetry group of order $N!\times N!\times 2$
    arising from:
    \begin{itemize}[nosep]
      \item $N!$ row permutations,
      \item $N!$ column permutations, and
      \item reflection in the main diagonal (transposition).
    \end{itemize}
    These symmetries can be exploited through symmetry-breaking
    constraints to drastically reduce the search
    space~\cite{gent2000symmetry}.

  \item \textbf{Objective (hardest-puzzle variant).}\;
    Among all feasible click matrices for all solvable targets, find the
    target whose minimum-click representative maximises
    $\sum_{i,j} x_{i,j}$.
\end{enumerate}

%% =====================================================================
\section{Integer Linear Programming (ILP) Model}\label{sec:ilp}

We present the ILP formulation following
Chlond~\cite{chlond2006alien}.

\subsection{Decision variables}

\begin{itemize}[nosep]
  \item $x_{i,j} \in \{0, 1, \dots, c-1\}$ \quad for
        $i,j \in \{1,\dots,N\}$: number of clicks on cell $(i,j)$.
  \item $q_{r,k} \in \Z_{\ge 0}$: auxiliary integer variable used to
        linearise the modulo constraint for cell $(r,k)$.
\end{itemize}

\subsection{Constraints}

The modular constraint~\eqref{eq:feasibility} is linearised by
introducing the auxiliary variables $q_{r,k}$:
\begin{equation}\label{eq:ilp-constraint}
  \sum_{j=1}^{N} x_{r,j} + \sum_{i=1}^{N} x_{i,k} - x_{r,k}
  - c\, q_{r,k} = t_{r,k},
  \qquad \forall\, r,k \in \{1,\dots,N\}.
\end{equation}

\noindent\textbf{Variable bounds:}
\begin{align}
  0 &\le x_{i,j} \le c - 1,
      &\forall\, i,j \in \{1,\dots,N\}, \label{eq:x-bounds}\\[4pt]
  0 &\le q_{r,k} \le \left\lfloor
        \frac{(2N-1)(c-1)}{c} \right\rfloor,
      &\forall\, r,k \in \{1,\dots,N\}. \label{eq:q-bounds}
\end{align}
The upper bound on $q_{r,k}$ follows from the fact that
$\sum_j x_{r,j} + \sum_i x_{i,k} - x_{r,k} \le (2N-1)(c-1)$.

\subsection{Objective function}

\paragraph{Variant 1: Solving a given target.}
If the goal is simply to find \emph{any} feasible click matrix for a
given target~$T$, no objective is needed (feasibility problem).

\paragraph{Variant 2: Minimum total clicks for a given target.}
\begin{equation}\label{eq:obj-min}
  \min \sum_{i=1}^{N}\sum_{j=1}^{N} x_{i,j}
  \quad \text{subject to } \eqref{eq:ilp-constraint}\text{--}\eqref{eq:q-bounds}.
\end{equation}

\paragraph{Variant 3: Hardest puzzle (max--min).}
This requires a bi-level formulation.  One approach is to maximise
over all solvable targets~$T$:
\begin{equation}\label{eq:obj-maxmin}
  \max_{T}\; \min_{X}\; \sum_{i,j} x_{i,j}
  \quad \text{s.t.\ } X \text{ solves } T.
\end{equation}
Gent et al.\ solved this via constraint programming with symmetry
breaking~\cite{gent2000symmetry}; Chlond tackled it with
ILP~\cite{chlond2006alien}.

\subsection{Symmetry-breaking constraints (optional)}

To exploit the $N!\times N!\times 2$ symmetries, one can add:
\begin{align}
  \sum_{j=1}^{N} x_{1,j} &\le \sum_{j=1}^{N} x_{2,j}
                          \le \cdots
                          \le \sum_{j=1}^{N} x_{N,j},
                          \label{eq:row-sym}\\[4pt]
  \sum_{i=1}^{N} x_{i,1} &\le \sum_{i=1}^{N} x_{i,2}
                          \le \cdots
                          \le \sum_{i=1}^{N} x_{i,N},
                          \label{eq:col-sym}\\[4pt]
  x_{1,2} &\le x_{2,1}. \label{eq:diag-sym}
\end{align}
These break row permutation, column permutation, and diagonal
reflection symmetries, respectively~\cite{gent2000symmetry}.

%% =====================================================================
\section{SAT / Pseudo-Boolean Model}\label{sec:sat}

We now encode the Alien Tiles problem as a propositional satisfiability
(SAT) instance.

\subsection{Binary encoding of variables}

Each click variable $x_{i,j}$ takes values in $\{0,\ldots,c-1\}$ and
is encoded using $b = \lceil\log_2 c\rceil$ Boolean variables:
\[
  x_{i,j} = \sum_{\ell=0}^{b-1} 2^\ell\, p_{i,j,\ell},
  \qquad p_{i,j,\ell} \in \{0,1\}.
\]
If $c$ is not a power of~2, additional \emph{at-most} clauses exclude
the unused values $c, c+1, \ldots, 2^b - 1$.

\subsection{Linearised sum variables}

Define the sum for each cell $(r,k)$ as:
\[
  \sigma_{r,k} = \sum_{j=1}^{N} x_{r,j}
                 + \sum_{i=1}^{N} x_{i,k}
                 - x_{r,k}.
\]
This sum can range from $0$ to $(2N-1)(c-1)$.  Represent $\sigma_{r,k}$
in binary using $B = \lceil\log_2((2N-1)(c-1)+1)\rceil$ bits.  The
addition can be implemented via a \emph{binary adder network} (e.g.,
ripple-carry or parallel-prefix adders) translated into CNF clauses
using the Tseitin transformation.

\subsection{Modulo constraint in CNF}

The constraint $\sigma_{r,k} \bmod c = t_{r,k}$ can be encoded as:
\[
  \sigma_{r,k} - c\,q_{r,k} = t_{r,k},
\]
where $q_{r,k}$ is an auxiliary integer variable encoded in binary.
Alternatively, one can enumerate all valid assignments:
\[
  \sigma_{r,k} \in \{t_{r,k},\; t_{r,k} + c,\; t_{r,k} + 2c,\;
                     \dots\},
\]
and encode this as a disjunction of equality constraints, each expanded
into CNF.

\subsection{Complexity summary}

\begin{center}
\begin{tabular}{@{}lcc@{}}
\toprule
\textbf{Component} & \textbf{Boolean variables} & \textbf{Clauses} \\
\midrule
Click variables $p_{i,j,\ell}$
  & $N^2 \lceil\log_2 c\rceil$ & --- \\
Sum auxiliary bits
  & $O(N^2 \log(Nc))$ & $O(N^3 \log(Nc))$ \\
Modulo constraints
  & $O(N^2 \log(Nc))$ & $O(N^2 \cdot Nc / c)$ \\
Value exclusion (if $c \ne 2^b$)
  & --- & $O(N^2)$ \\
\bottomrule
\end{tabular}
\end{center}

%% =====================================================================
\section{SAT-Based Optimisation Approaches}\label{sec:optim}

The feasibility encoding of \Cref{sec:sat} determines whether a given
target~$T$ is solvable.  For the \emph{minimisation} variant
($\min \sum_{i,j} x_{i,j}$), we must couple the feasibility formula
with an objective.  Three principal paradigms exist: non-incremental
SAT, incremental SAT, and MaxSAT / pseudo-Boolean optimisation.

\paragraph{Unary expansion of the objective.}
To link the click values to a cardinality objective, we introduce
\emph{unit-contribution literals}.  For each cell $(i,j)$ and each
$v \in \{1,\dots,c{-}1\}$, define
\begin{equation}\label{eq:unit}
  u_{i,j,v} = 1 \;\Longleftrightarrow\; x_{i,j} \ge v.
\end{equation}
Then $\text{total}(X) = \sum_{i,j}\sum_{v=1}^{c-1} u_{i,j,v}$, and
minimising total clicks becomes minimising the number of true
$u$-literals.  The channelling between $u_{i,j,v}$ and the click
variables $d_{i,j,w}$ (``$x_{i,j}=w$'') is:
\[
  u_{i,j,v} \;\Longleftrightarrow\;
  \bigvee_{w=v}^{c-1} d_{i,j,w}.
\]
This yields $n = N^2(c{-}1)$ unit literals in total.

% ---------------------------------------------------------------------
\subsection{Non-incremental SAT (multiple independent calls)}%
\label{sec:optim-noninc}

A sequence of \emph{independent} SAT instances is solved, each with a
different cardinality bound.  A fresh solver is created for every call,
so learned clauses are \emph{not} reused.

\subsubsection{Binary search}

Search for the minimum $K$ such that $\mathcal{F} \wedge
\textsc{AtMost}(\{u_{i,j,v}\}, K)$ is satisfiable, where
$\mathcal{F}$ denotes the feasibility formula:

\begin{algorithmic}[1]
\State $\ell \gets 0$;\; $h \gets N^2(c-1)$;\;
       $X^* \gets \textsc{None}$
\While{$\ell \le h$}
  \State $m \gets \lfloor (\ell+h)/2 \rfloor$
  \State Build fresh CNF: $\mathcal{F}
         \wedge \textsc{AtMost}(\{u\}, m)$
  \If{\textsc{Sat}(\,)}
    \State $X^* \gets \textsc{Extract}()$;\;
           $h \gets m - 1$
  \Else\; $\ell \gets m + 1$
  \EndIf
\EndWhile
\State \Return $X^*$
  \Comment{optimal in
  $\lceil\log_2(N^2(c{-}1))\rceil$ SAT calls}
\end{algorithmic}

\medskip
The \textsc{AtMost} constraint can be encoded via:
\begin{itemize}[nosep]
  \item \textbf{Sequential counter}~\cite{sinz2005towards}:
    $O(n\!\cdot\! K)$ auxiliary variables and clauses; compact with
    strong unit propagation.
  \item \textbf{Totalizer}~\cite{bailleux2003efficient}:
    $O(n\log n)$ variables, $O(n\log^2 n)$ clauses; often faster for
    large~$n$.
  \item \textbf{Cardinality networks}~\cite{asin2011cardinality}:
    sorting-network based; balances formula size and propagation
    strength.
\end{itemize}

\subsubsection{Linear search (top-down)}

Find any feasible solution~$X$, record its total~$K$, then solve with
bound~$K{-}1$, repeating until unsatisfiable.  This needs
$K_{\text{first}} - K^* + 1$ calls in the worst case and is effective
when a good initial solution is found early.

% ---------------------------------------------------------------------
\subsection{Incremental SAT}\label{sec:optim-inc}

A \emph{single} solver instance is kept alive across iterations.
Learned clauses are retained, making subsequent calls faster.

\subsubsection{Assumption-based bound control}

The cardinality encoding is built \emph{once}.  An \emph{incremental
totalizer}~\cite{martins2014incremental} produces output literals
$o_1, o_2, \dots, o_n$ where $o_k$ is true if and only if at least~$k$
of the unit literals are true.  To assert ``at most~$K$'', the solver
is called with the \emph{assumption} $\neg o_{K+1}$:

\begin{algorithmic}[1]
\State \textbf{solver} $\gets$
       \textsc{IncrementalSolver}($\mathcal{F}$)
\State Build totalizer on $\{u_{i,j,v}\}$
       $\;\to\;$ outputs $o_1,\dots,o_n$
\State Add totalizer clauses to \textbf{solver}
\State $\ell \gets 0$;\; $h \gets n$;\;
       $X^* \gets \textsc{None}$
\While{$\ell \le h$}
  \State $m \gets \lfloor(\ell+h)/2\rfloor$
  \If{$\textbf{solver}.\textsc{Solve}(
      \text{assumptions}=\{\neg o_{m+1}\})$}
    \State $X^* \gets \textsc{Extract}()$;\;
           $h \gets m - 1$
  \Else\; $\ell \gets m + 1$
  \EndIf
\EndWhile
\State \Return $X^*$
\end{algorithmic}

\medskip
\noindent Key advantages over non-incremental search:
\begin{itemize}[nosep]
  \item The totalizer is built and added to the solver \emph{once}
        (no re-encoding per call).
  \item Learned clauses from earlier iterations remain valid and
        accelerate later calls --- typically $2$--$10\times$ faster.
  \item Changing the bound only modifies the \emph{assumptions}, not
        the clause database.
\end{itemize}

% ---------------------------------------------------------------------
\subsection{MaxSAT and pseudo-Boolean optimisation}%
\label{sec:optim-maxsat}

Instead of iterating over bounds externally, the objective is encoded
\emph{inside} the formula and the solver handles optimisation
internally.

\subsubsection{Partial weighted MaxSAT}

The formula is split into \emph{hard} clauses (must be satisfied) and
\emph{soft} clauses (satisfy as many as possible):
\begin{itemize}[nosep]
  \item \textbf{Hard clauses:} all feasibility constraints
        $\mathcal{F}$ (exactly-one, running-sum, target assertion).
  \item \textbf{Soft clauses:} for each unit literal $u_{i,j,v}$, add
        the unit clause $(\neg u_{i,j,v})$ with weight~$1$.
\end{itemize}
Minimising total clicks is equivalent to \emph{maximising} the number of
satisfied soft clauses (i.e., maximising the literals $\neg u_{i,j,v}$
that are true).

Modern MaxSAT solvers employ several internal strategies:
\begin{itemize}[nosep]
  \item \textbf{Core-guided algorithms} (e.g., OLL~\cite{morgado2014iterative},
        MSU3~\cite{martins2014open}):
        extract unsatisfiable cores of soft clauses, add relaxation
        variables, and iterate.  Particularly effective when the optimum
        is far from the worst case.
  \item \textbf{Linear SAT--UNSAT search:} internally equivalent to
        \Cref{sec:optim-noninc} but with learned-clause reuse.
  \item \textbf{Stratification:} partition soft clauses by weight and
        solve layer by layer; useful for weighted instances.
\end{itemize}

\subsubsection{Pseudo-Boolean optimisation (PBO)}

A PBO solver (e.g., RoundingSat) minimises a linear pseudo-Boolean
objective directly:
\begin{equation}\label{eq:pbo}
  \min \sum_{i,j}\sum_{v=1}^{c-1} u_{i,j,v}
  \qquad\text{subject to } \mathcal{F}.
\end{equation}
PBO solvers use the \emph{cutting-planes} proof system, which is
strictly stronger than the resolution system underlying CDCL SAT
solvers, and can therefore produce shorter proofs of
optimality for certain structured instances.

\subsubsection{Weighted variant (binary encoding)}

If binary encoding is used for $x_{i,j}$ instead of one-hot, the
objective is a \emph{weighted} sum
$\text{total} = \sum_{i,j}\sum_{\ell=0}^{b-1} 2^\ell\, p_{i,j,\ell}$.
Each bit contributes a soft clause $(\neg p_{i,j,\ell})$ with weight
$2^\ell$, yielding a \emph{weighted} partial MaxSAT instance.
Core-guided solvers handle weights via weight-aware core extraction
(e.g., WPM1 with weight splitting).

% ---------------------------------------------------------------------
\subsection{Comparison of optimisation approaches}

\Cref{tab:optim-cmp} summarises the key trade-offs.

\begin{table}[ht]
\centering
\caption{Comparison of SAT-based optimisation strategies for Variant~2.}
\label{tab:optim-cmp}
\begin{tabular}{@{}lccc@{}}
\toprule
\textbf{Criterion}
  & \textbf{Non-incremental}
  & \textbf{Incremental}
  & \textbf{MaxSAT / PBO} \\
\midrule
SAT calls (external)
  & $O(\log(N^2 c))$
  & $O(\log(N^2 c))$
  & 1 \\
Learned-clause reuse
  & \textsf{no}
  & \textsf{yes}
  & \textsf{yes} \\
Cardinality encoder rebuilt
  & each call
  & once
  & once / internal \\
Implementation effort
  & low
  & medium
  & low \\
Optimality guarantee
  & binary search
  & binary search
  & solver-internal \\
\bottomrule
\end{tabular}
\end{table}

%% =====================================================================
\section{Datasets and Benchmarks}\label{sec:dataset}

\subsection{CSPLib instances}

The canonical dataset for the Alien Tiles problem is hosted at the
CSPLib repository~\cite{csplib027}:
\begin{center}
\url{https://www.csplib.org/Problems/prob027/}
\end{center}

CSPLib provides:
\begin{itemize}[nosep]
  \item An \textsc{Essence} model for single-pass solving
        (\texttt{prob027-alientiles-singlepass.essence}).
  \item A multi-pass model using equivalence to determine minimality
        (\texttt{prob027-alientiles-multipass.zip}), based on the method
        of Gent, Linton, and Smith~\cite{gent2000symmetry}.
  \item Known results: for $N=4,\, c=3$, the hardest instance requires
        $10$ clicks, and there are $19$ distinct hardest instances up to
        symmetry~\cite{gent2000symmetry}.
\end{itemize}

\subsection{Instance parameters}

\Cref{tab:instances} summarises selected instance families.

\begin{table}[ht]
\centering
\caption{Selected instance families for the Alien Tiles problem.}
\label{tab:instances}
\begin{tabular}{@{}ccccl@{}}
\toprule
$N$ & $c$ & \# Variables & Known optimum & Source \\
\midrule
4 & 3 & 16 & 10 clicks & Gent et al.~\cite{gent2000symmetry} \\
$N$ & $N{-}1$ & $N^2$ & open & CSPLib~\cite{csplib027} \\
7 & 4 & 49 & open & Chlond~\cite{chlond2006alien} \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Generating test instances.}
For $N\times N$ with $c = N-1$ colours, clicking 1~time in each cell of
row~1 and $(N-2)$~times in each cell of row~2 generates a null pattern,
providing a convenient test instance family~\cite{csplib027}.

\subsection{Computational results from the literature}

Gent, Linton, and Smith~\cite{gent2000symmetry} solved the $N{=}4,\,
c{=}3$ instance using ILOG Solver with GAP-generated symmetry-breaking:
\begin{itemize}[nosep]
  \item Without symmetry breaking: 866 fails, 31\,s to find optimum;
        57\,664 fails, 2\,324\,s to prove optimality.
  \item With symmetry breaking: 116 fails, 18.6\,s;
        499 fails, 56.6\,s to prove optimality.
\end{itemize}

Chlond~\cite{chlond2006alien} formulated the $7\times 7,\, c{=}4$
puzzle as an ILP and noted that transforming the entire grid to a single
colour is particularly challenging, even with symmetry elimination.

%% =====================================================================
\section{Conclusion}\label{sec:conclusion}

The Alien Tiles problem is a rich combinatorial puzzle with connections
to coding theory, linear algebra over finite fields, and symmetry
exploitation.  We have presented formal ILP and SAT encodings that can
be used to solve instances and to search for the hardest puzzles.  The
CSPLib repository provides standard models and benchmarks for further
research.

%% =====================================================================
\bibliographystyle{plain}
\bibliography{references/references}

\end{document}
